A partisan affair: Mapping patronage in municipal bureaucracies of Brazil

### 2021-01-14 paper/__latexindent_temp.tex
generate intro

In this paper, I focus on party membership as a determinant of not only whether or not individuals enter into public sector employment, but the kind of employment you receive. In particular, using a rich panel dataset of all public sector workers, I identify the party membership of all municipal bureaucrats in Brazil to provide a novel set of empirical findings: 1) the pre-bureaucracy characteristics of public sector workers, 2)

In this paper, I leverage micro-level partisan information with employment data to assess the degree to which there are differential compensation structures and differences in observables among partisan and non-partisan members of the bureaucracy. The main finding is that patronage is an elite game, benefitting the wealthiest formal sector workers and allocating them into the best-paying jobs in the public sector. The benefits of public sector employment accrue to a wealthier class of workers, in particular those already in profitable positions in the private sector. Additionally, wealthy party members accumulate jobs at higher levels of government such as executive leadership and administration, which are better compensated. In contrast, less profitable offices in frontline services experience less exposure to partisan members.

Additionally, party members are concentrated in tenured positions. Over X\% of career bureaucrats are affiliated to a party. Party members tend to stay in the bureaucracy for longer, and receive a higher compensation than their non-partisan counterparts.

Patronage in the allocation of public sector jobs

>>>>

In this paper, I focus on party membership as a determinant of not only whether or not individuals enter into public sector employment, but the type of employment they receive. In particular, using a rich panel data set of all public sector workers, I identify the party membership of all municipal bureaucrats in Brazil to provide a novel set of empirical findings: 1) the pre-bureaucracy characteristics of public sector workers, 2) employment trajectories within the bureaucracy and 3) post-public sector employment of party members. This complete revolving-door of party and non-party members provide a unique frame-by-frame evolution of the employment trajectories of pre- and post-bureaucrats in a developing world context.

The main finding is that patronage primarily benefits a local economic elite, accruing to the richest formal sector workers and allocating them into the best-paying jobs in the municipal bureaucracy. These higher compensation structures are not commensurate to skills on a set of observable qualifications, such as education level and work experience, suggesting that these benefits accrue from channels other than individual skills. Additionally, party members accumulate jobs at higher levels of government such as executive leadership and administration, which are better compensated, as well as benefiting from longer income streams due to privileged access to tenured contracts.

These findings suggest that, overall, patronage does not necessarily accrue to poor voters, contrary to theoretical and empirical findings in studies of clientelism in the developing world \citep{stokes2013brokers}. A different logic seems to be at play. Patronage to party loyalists can be used strategically as a mean of securing control over the bureaucracy -- through tenured contracts -- as well as securing buy-in from wealthy patrons or notables in the local economy, consistent with empirical findings by \citet{colonnelli2018patronage}. Patronage therefore can serve a crucial role in securing access to economic resources through wealthy patrons, which in turn can be allocated to finance campaigns and other exercises in party building.
